Program Transformations for miniKanren

### 2020-07-03 papers/aplas/evaluation.tex
Evaluation draft

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
    \begin{lstlisting}[label={doubleApp}, caption={Concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res =
    fresh (ts) (
      append$^o$ xs ys ts /\ append$^o$ ts zs res)

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
    \begin{lstlisting}[label={doubleApp:etalon}, caption={Etalon implementation of concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res = conde [
    (xs === [] /\ append$^o$ ys zs res);
    fresh (h t r) (
      xs === h % t /\
      res === h % r /\
      doubleAppend$^o$ t ys zs r)]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
\begin{lstlisting}[label={cpd:maxandlength}, caption={Maximum element and length of the list}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = max$^o$ xs m /\ length$^o$ xs l
  let rec length$^o$ xs l = conde [
    (xs === [] /\ l === zero);
    (fresh (h t m) (
      xs === h % t /\ l === succ m /\ length$^o$ t m))]
  let max$^o$ xs m = max$_1^o$ xs zero m
  let rec max$_1^o$ xs n m = conde [
    (xs === [] /\ m === n);
    (fresh (h t) (
      (xs === h % t) /\
      (conde [
        (le$^o$ h n ^true /\ max$_1^o$ t n m);
        (gt$^o$ h n ^true /\ max$_1^o$ t h m)])))]
  let rec le$^o$ x y b = conde [
    (x === zero /\ b === ^true);
    (fresh (x$_1$) (
      x === succ x$_1$ /\ y === zero /\ b === ^false));
    (fresh (x$_1$ y$_1$) (
      x === succ x$_1$ /\ y === succ y$_1$ /\ le$^o$ x$_1$ y$_1$ b))]
  let rec gt$^o$ x y b = conde [
    (x === zero /\ b === ^false);
    (fresh (x$_1$) (
      x === succ x$_1$ /\ y === zero /\ b === ^false));
    (fresh (x$_1$ y$_1$) (
      x === succ x$_1$ /\ y === succ y$_1$ /\ gt$^o$ x$_1$ y$_1$ b))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
\begin{lstlisting}[label={etalon:maxandlength}, caption={Etalon implementation of maxlengtho}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = maxLength$_1^o$ xs m zero l
  let rec maxLength$_1^o$ xs m n l = conde [
    (xs === [] /\ m === n /\ l === zero);
    (fresh (h t l$_1$)
       (xs === h % t) /\
       (l === succ l$_1$) /\
       (conde [
         (le$^o$ h n ^true /\ maxLength$_1^o$ t m n l);
         (gt$^o$ h n ^true /\ maxLength$_1^o$ t m h l)]))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.8\textwidth}
    \begin{lstlisting}[label={eval:last}, caption={Evaluator of formulas with boolean operation last}, captionpos=b, frame=tb]
  let rec eval$^o$ subst fm res = conde [
    fresh (x y z v w) (
      (fm === var v /\ elem$^o$ subst v res);
      (fm === conj x y /\ eval$^o$ st x v /\ eval$^o$ st y w /\ and$^o$ v w res);
      (fm === disj x y /\ eval$^o$ st x v /\ eval$^o$ st y w /\ or$^o$ v w res);
      (fm === neg x /\ eval$^o$ st x v /\ not$^o$ v res))]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.85\textwidth}
    \begin{lstlisting}[label={eval:fst}, caption={Evaluator of formulas with boolean operation second}, captionpos=b, frame=tb]
  let rec eval$^o$ subst fm res = conde [
    fresh (x y z v w) (
      (fm === var v /\ elem$^o$ subst v res);
      (fm === conj x y /\ and$^o$ v w res /\ eval$^o$ st x v /\ eval$^o$ st y w);
      (fm === disj x y /\ or$^o$ v w res /\ eval$^o$ st x v /\ eval$^o$ st y w);
      (fm === neg x /\ not$^o$ v res /\ eval$^o$ st x v))]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
    \begin{lstlisting}[label={not:nando}, caption={Implementation of boolean operation via \lstinline{nand}}, captionpos=b, frame=tb]
  let not$^o$ x y = nand$^o$ x x y

@@

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c||c}
                   & \multicolumn{3}{c}{unify} & isPath \\
  \hline
                   & first  & second & third   &        \\
  \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$ (oum?) & $>30s$ (oum?) & $23.743s$ \\
  \hline
  CPD              & $0.081*10^{-4}s$ & $0.017s$ & $0.343s$  & $2.999s$  \\
  \hline
  Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
  \hline
  Ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
  \hline
  Branching        & $0.205*10^{-4}s$ & $0.101s$ & $>30s$ (oum?) & N/A \\
  \hline
  \end{tabular}

@@

This relation implements a typechecker for a tiny language. Being executed in the backward direction it serves as a generator of terms of the given type.

Its abstract syntax is the following.

@@

We use de Bruijn indices for variables.

Typing rules are the following

@@

Implementation 1 is a straightforward implementation in a relational programming language.

Implementation 2 is obtained by translating into \mk{} an implementation of the typing rules \ocaml{}.

We measured the time needed to generate 100 formulas (averaged over 10 iterations).

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\
  \hline\hline
  Original    & 0.057s & 0.464s\\
  \hline
  Non-CPD     & 0.154s & 0.448s \\
  \hline
  Ecce        & 2.543s  & 1.518s \\
  \end{tabular}

@@

>>>>

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.6\textwidth}
    \begin{lstlisting}[label={doubleApp}, caption={Concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res =
    fresh (ts) (
      append$^o$ xs ys ts /\ append$^o$ ts zs res)

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.6\textwidth}
    \begin{lstlisting}[label={doubleApp:etalon}, caption={Etalon implementation of concatenation of three lists}, captionpos=b, frame=tb]
  let doubleAppend$^o$ xs ys zs res = conde [
    (xs === [] /\ append$^o$ ys zs res);
    fresh (h t r) (
      xs === h % t /\
      res === h % r /\
      doubleAppend$^o$ t ys zs r)]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.85\textwidth}
\begin{lstlisting}[label={cpd:maxandlength}, caption={Maximum element and length of the list}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = max$^o$ xs m /\ length$^o$ xs l
  let rec length$^o$ xs l = conde [
    (xs === [] /\ l === zero);
    (fresh (h t m) (
      xs === h % t /\ l === succ m /\ length$^o$ t m))]
  let max$^o$ xs m = max$_1^o$ xs zero m
  let rec max$_1^o$ xs n m = conde [
    (xs === [] /\ m === n);
    (fresh (h t) (
      (xs === h % t) /\
      (conde [
        (le$^o$ h n ^true /\ max$_1^o$ t n m);
        (gt$^o$ h n ^true /\ max$_1^o$ t h m)])))]
  let rec le$^o$ x y b = conde [
    (x === zero /\ b === ^true);
    (fresh (x$_1$) (x === succ x$_1$ /\ y === zero /\ b === ^false));
    (fresh (x$_1$ y$_1$) (x === succ x$_1$ /\ y === succ y$_1$ /\ le$^o$ x$_1$ y$_1$ b))]
  let rec gt$^o$ x y b = conde [
    (x === zero /\ b === ^false);
    (fresh (x$_1$) (x === succ x$_1$ /\ y === zero /\ b === ^false));
    (fresh (x$_1$ y$_1$) (x === succ x$_1$ /\ y === succ y$_1$ /\ gt$^o$ x$_1$ y$_1$ b))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
\begin{lstlisting}[label={etalon:maxandlength}, caption={Etalon implementation of maxlengtho}, captionpos=b, frame=tb]
  let maxLength$^o$ xs m l = maxLength$_1^o$ xs m zero l
  let rec maxLength$_1^o$ xs m n l = conde [
    (xs === [] /\ m === n /\ l === zero);
    (fresh (h t l$_1$)
       (xs === h % t) /\
       (l === succ l$_1$) /\
       (conde [
         (le$^o$ h n ^true /\ maxLength$_1^o$ t m n l);
         (gt$^o$ h n ^true /\ maxLength$_1^o$ t m h l)]))]
  \end{lstlisting}
\end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.95\textwidth}
    \begin{lstlisting}[label={eval:last}, caption={Evaluator of formulas with boolean operation last}, captionpos=b, frame=tb]
  let rec eval$^o$ subst fm res = conde [
    fresh (x y z v w) (
      (fm === var v /\ elem$^o$ subst v res);
      (fm === conj x y /\ eval$^o$ st x v /\ eval$^o$ st y w /\ and$^o$ v w res);
      (fm === disj x y /\ eval$^o$ st x v /\ eval$^o$ st y w /\ or$^o$ v w res);
      (fm === neg x /\ eval$^o$ st x v /\ not$^o$ v res))]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.95\textwidth}
    \begin{lstlisting}[label={eval:fst}, caption={Evaluator of formulas with boolean operation second}, captionpos=b, frame=tb]
  let rec eval$^o$ subst fm res = conde [
    fresh (x y z v w) (
      (fm === var v /\ elem$^o$ subst v res);
      (fm === conj x y /\ and$^o$ v w res /\ eval$^o$ st x v /\ eval$^o$ st y w);
      (fm === disj x y /\ or$^o$ v w res /\ eval$^o$ st x v /\ eval$^o$ st y w);
      (fm === neg x /\ not$^o$ v res /\ eval$^o$ st x v))]
    \end{lstlisting}
  \end{minipage}
\end{figure*}

@@

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.85\textwidth}
    \begin{lstlisting}[label={not:nando}, caption={Implementation of boolean operation via \lstinline{nand}}, captionpos=b, frame=tb]
  let not$^o$ x y = nand$^o$ x x y

@@

\begin{table}
  \centering
  \begin{tabular}{c||c|c|c||c}
                   & \multicolumn{3}{c||}{unify} & isPath \\
  \hline
                   & first  & second & third     &        \\
  \hline\hline
  Original         & $0.247*10^{-4}s$ & $>30s$ (oum?) & $>30s$ (oum?) & $23.743s$ \\
  \hline
  CPD              & $0.081*10^{-4}s$ & $0.017s$ & $0.343s$  & $2.999s$  \\
  \hline
  Non-CPD          & $0.084*10^{-4}s$ & $0.021s$ & $2.198s$  & $1.786s$  \\
  \hline
  Ecce             & $0.128*10^{-4}s$ & $0.026s$ & $1.340s$  & $1.292s$ \\
  \hline
  Branching        & $0.205*10^{-4}s$ & $0.101s$ & $>30s$ (oum?) & N/A \\
  \hline
  \end{tabular}

@@

This relation implements a typechecker for a tiny language.
Being executed in the backward direction it serves as a generator of terms of the given type.
The abstract syntax of the language is presented below.
The variables are represented with de Bruijn indices, thus let-binding does not specify which variable is being bound.

@@

The typing rules are straightforward and are presented below.
Boolean and integer constants have the corresponding types regardless of the environment.
Only terms of type integer can be summed up, multiplied or checked for being less or equal.
Any terms of the same type can be checked for equality.
If-then-else expression typechecks only if its condition is of type boolean, while both then- and else-branches have the same type.
An environment $\Gamma$ is an ordered list, in which the $i$-th element is the type of the variable with the $i$-th de Bruijn index.
To typecheck a let-binding, first, the term being bound is typechecked and is added in the beginning of the environment $\Gamma$, and then the body is typechecked in the context of the new environment.
Typechecking a variable with the index $i$ boils down to getting a $i$-th element of the list.

@@

We compared two possible implementations of these typing rules.
The first one is obtained by unnesting of the functional program as described in~\cite{lozov2019relational}.
The second program is hand-written in \oc{}.
Each implementation has been transformed with non-conjunctive partial deducer and by Ecce.

We measured the time needed to generate 100 terms of type integer (averaged over 10 iterations).
Ecce significantly worsens the execution time for both implementations.
Non-conjunctive partial deduction improves the first implementation a little bit, while worsening the performance of the second implementation 3 times.

\begin{table}
  \centering
  \begin{tabular}{c||c||c}
              & Implementation 1 & Implementation 2 \\ \hline\hline
  Original    & 0.464s           & 0.057s           \\ \hline
  Non-CPD     & 0.448s           & 0.154s           \\ \hline
  Ecce        & 1.518s           & 2.543s           \\
  \end{tabular}

@@

\todo{why}
